Bumble bees and wildflowers along a mountain slope exhibit nonlinear patterns of abundance and turnover, punctuated by the tree line ecotone

### 2021-03-03 writing/ms_abundance_turnover.Rmd
update

---
title: Bumble bees exhibit strong but flexible foraging biases through temporal and
  elevational gradients of floristic turnover
author:
- Douglas B. Sponsler
- Fabrice Requier
- Katharina Kallnik
- Alice Claßen
- A. Fabienne Maihoff
- Ingolf Steffan-Dewenter
bibliography: refs.bib
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document:
    fig_caption: yes
---

@@

Bumble bees are exceptionally versatile foragers whose ability to solve the puzzles posed by complex flower morphologies [@Plowright1984-lc], or indeed by human investigators [@Alem2016-rn; @Loukola2017-bt], renders almost any flower potentially accessible. Even when a gross mismatch between tongue length and corolla depth makes the nectar of a given flower inaccessible by ordinary means, some bumble bee species will resort readily to biting through the base of the flower to “rob” nectar, incidentally making the nectar available to other prospective foragers as well. Flower handling skills are, moreover, socially transmitted, amplifying the plasticity of individual behavior [@Leadbeater2008-rh; @Loukola2017-bt]. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the searching/learning/handling costs associated with exploiting it. At an evolutionary level, the innate biases of bumble bees for certain floral traits likely reflect historic associations between floral traits and floral rewards [@Raine2007-nq]. These innate biases, however, can be modulated or even nullified by patterns of reward learned during foraging [@Kugler1943-uy; @Maharaj2019-wm], which depend not only on intrinsic floral properties but also ecological modulators, such as the spatial distribution of floral patches or competition with other bumble bees [@Ishii2013-sm; @Plowright1984-lc; @Sowig1989-ld].

The topic of floral bias in bumble bees has been approached via both laboratory experimentation [e.g. @Kugler1943-uy; @Raine2007-nq] and observational field study [e.g. @Heinrich1976-wc; @Goulson2008-zy]. The former approach enables stronger causal inference with respect to the mechanisms of floral choice, but it suffers from all the usual limitations of biological realism associated with laboratory conditions. Observational field studies, however, tend to be anecdotal in scope, typically describing bumble bee foraging patterns in particular localities, over short time frames, and with limited floristic data to enable a distinction between true foraging bias and the neutral effects of floral abundance. The steep elevation gradients of mountain slopes provide an opportunity for more powerful observational study of bumble foraging because they generate broad gradients of environmental variation while controlling for confounding effects of geographic variation, such as genotypic differentiation within bumble bee species. When bumble bee visitation in such environments is sampled with sufficient spatial and temporal replication and complemented with quantitative floral surveying, true bias can be distinguished from the neutral effects of abundance, and variation in bias can be explored along elevational and temporal dimensions of floristic turnover.

In the present study, we focus on the question of bumble bee foraging bias with respect to floral morphology, and the lability of morphological biases across gradients of floristic variation. At the outset, we emphasize the use of the term "bias" in favor of the term "preference", since the latter is laden with more assumptions and its inference under field conditions is problematic [@Williams2005-yp]. By "bias", we mean simply an empirical disproportionality between the relative abundance of a given floral taxon or group and its proportional visitation. The exact mechanism of this disproportionality --- whether an arbitrary "preference" on the part of the foraging bee, an unmeasured decoupling of relative abundance and rate of reward, an exogenous effect of competition, etc. --- is beyond the scope of our discussion, and probably best studied under laboratory conditions where strict factorial control is possible. We focus instead on the empirical phenomenon of bias, its variation, and its morphological correlations.

The meaning of "morphology", however, is hardly more clear-cut than that of bias. Bumble bees and flowers vary in an array of interrelated morphological parameters, both within and between species, and the selection of relevant and measurable morphological metrics is nontrivial. For bumble bees, morphological variation consists principally (though not exclusively) of variation in tongue length. Tongue length in bumble bees corresponds functionally to corolla depth in flowers, and there is a well-established (though imperfect) association between tongue length and floral choice [e.g. @Inouye1980-mv]. Importantly, though, accessing floral reward involves more than just a match between tongue length and corolla depth. Such matching is entirely bypassed by nectar robbers, but even in the case of "legitimate" foraging, access to reward involves learning to manipulate floral structures that, in some floral morphotypes, are challenging irrespective of tongue length [@Laverty1994-hm]. For example, the characteristic Papilionaceous floral morphology of legumes requires a bee to separate tightly appressed keel petals to access anthers and nectaries [@Westerkamp1997-hs], and the narrowly campanulate flowers of many Ericaceous flora pose a significant barrier to large-bodied visitors like bumble bees [@Wallace1977-kz]. Rather than considering various floral morphological traits independently, we rely on the synthesis of @Kugler1970-gb, who developed a typology consisting of ten primary morphotypes and nested subdivision of each. Kugler's typology was informed by laboratory studies of bumble bee flower choice [@Kugler1943-uy], and there is a precedent for analyzing bumble bee visitation in terms of Kugler's morphotypes [@Neumayer1999-il; @Schneller2014-ry].

First, we compare the overall rate of bumble bee visitation across floral morphotypes. Then, we parse visitation rates by bumble bee species and explore the functional relationship between floral morphotype and bumble bee tongue length. These biases, together with the background effects of relative abundance, generate emergent patterns of diet (dis)similarity, which we present in the form of ordination analysis. Lastly, we analyze the lability of the morphological biases of individual bumble bee species through elevation and time, and compare these patterns of bias across bumble bee species and within floral morphtypes. The results of these analyses demonstrate both the strength of bumble bee floral bias, arising in part from morphological constraints, and the flexibility of bumble bee floral bias that enables them to forage adaptively in the dynamic floral communities of mountain landscapes.

>>>>

---
title: Bumble bees and wildflowers along a mountain slope exhibit nonlinear patterns of abundance and turnover, punctuated by the tree line ecotone
author:
- Douglas B. Sponsler
- Fabrice Requier
- Katharina Kallnik
- Alice Claßen
- A. Fabienne Maihoff
- Ingolf Steffan-Dewenter
bibliography: refs.bib
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document:
    fig_caption: yes
---

@@

Anthropogenic climate change is occurring at ecological rather than evolutionary time scales, forcing species to adapt within the constraints of gene pools shaped by cooler climates [@Visser2008-ct]. As temperatures rise, species are expected to be pushed to higher latitudes and higher elevations, and such effects are already measurable across a broad range of taxa [@Lenoir2015-tj].
